---
title: "minor: skull and manus analyses, MS edition"
author: "Maya Juman"
date: "September 3, 2021"
output: html_document
---

```{r}
library("ggplot2")
library("rnaturalearth")
library("rnaturalearthdata")
library("ggspatial")
```

```{r}
library(sf)
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
sites <- read.csv("minor_loc.csv")

map <- ggplot(data = world) + geom_sf() +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), panel.grid.major = element_line(colour = "transparent"), panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  coord_sf(xlim = c(97.5,119.5), ylim = c(-6.5,8), expand = FALSE) + xlab("Longitude") + ylab("Latitude") + 
  scale_shape_manual(values = c(22,24,16,18,21), 
                     guide = guide_legend(label.theme = element_text(angle = 0, 
                                                                     face = "italic", size=9), ncol=3)) + theme(legend.title = element_blank()) + 
  theme(legend.position = c(0.741,0.0865),
        legend.spacing.x = unit(0.5, 'mm'), legend.key.size = unit(5.4, 'mm'), legend.background = element_rect(fill = "transparent")) + annotation_scale(location = "bl", width_hint = 0.14) +
    annotation_north_arrow(location = "bl", which_north = "true", 
        pad_x = unit(0.3, "in"), pad_y = unit(0.25, "in"), 
        height = unit(1, "cm"),
        width = unit(1, "cm"),
        style = north_arrow_fancy_orienteering) + geom_point(data = sites, aes(x = Long, y = Lat, shape = Subspecies), size = 3, color = "black", fill = "white") + theme(axis.text.x = element_text(face="bold", color="black"), axis.text.y = element_text(face="bold", color="black")) + geom_point(aes(x=104.42, y=-0.49), colour="black", size = 0.6, shape = 3) + geom_point(aes(x=115.44, y=5.33), colour="white", size = 0.5, shape = 3) + geom_point(aes(x=116.93, y=7.27), colour="black", fill="white", size = 3, shape = 22) + geom_point(aes(x=116.93, y=7.27), colour="black", size = 0.4, shape = 3) + geom_point(aes(x=102.25, y=2.19), colour="black", size = 1.5, shape = 8) + geom_point(aes(x=102.43, y=-3.3), colour="black", size = 1.5, shape = 8)

map

#ggsave(map,file="Fig. 1.eps",width=200, height=120, units="mm")
#ggsave(map,file="Fig. 1.jpg",width=200, height=120, units="mm")
ggsave(map,file="Fig. 1.pdf",width=200, height=120, units="mm")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggfortify)
library(MASS)
library(klaR)
library(lme4)
library(car)
library(plyr)
library(gridExtra)

m <- c("grey60", "grey90", "black")
n <- c("grey80", "black")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#skull data
setwd("~/Desktop/minor/")
x <- read.csv("mskull.csv")
x[,11:32] <- log(x[,11:32])
x <- x[-c(51),]
table(x$Subspecies)
x <- x[-c(4,12,45),] #dump certain rows: 4,12,45 because heavy missing data, 51 because wrong species (not included in supp data)
x <- subset(x, select=-c(EPL,EB,Locality)) #remove certain columns
x$Sex[x$Sex == ""] <- "sex unknown"
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#hand data
setwd("~/Desktop/minor/")
y <- read.csv("minormanus.csv")
y[,19:56] <- log(y[,19:56])
y <- y[which(y$Age == "A"),] #only adults
y <- subset(y, select = -c(Age, Locality, TOT, HB, Tail, HF, ear, WT, PercentTL, X2MPL, X2DPW, X3MPL, X3MPW, X4MPL, X4DPW, X5MPL, X5MPW))
table(y$Subspecies)
y <- y[-1,]
y$Sex <- as.character(y$Sex)
y$Sex[y$Sex == "F"] <- "f"
y$Sex[y$Sex == "M"] <- "m"
y$Sex[y$Sex == ""] <- "sex unknown"
y$Sex[y$Sex == "?"] <- "sex unknown"
y$Sex <- as.factor(y$Sex)
```

##1. Geographic variation on Borneo

###a. Are mainland Bornean *T. m. caedis* distinct from mainland Bornean *T. m. minor*

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#hand
borneo2 <- y[y$Island == "Borneo",]
borneo2 <- na.omit(subset(borneo2, select = c(Inst, Cat., Subspecies, Island, region, lat, long, Sex, Side, X1ML, X1PPL, X1MD, X1PPD, X1DPD, X2PPL, X2MW, X5ML)))

borneo2$Subspecies <- droplevels(borneo2$Subspecies)
table(borneo2$Subspecies)

pc8 <- princomp(borneo2[,-c(1:9)], cor=TRUE)
for (i in 1:8) {
  pc8$loadings[,i] <- (pc8$loadings[,i] * pc8$sdev[i])
}
print(summary(pc8),digits=2,loadings=pc8$loadings,cutoff=0)
round(pc8$sdev^2,2)
borneo2$pc1 <- pc8[["scores"]][,1]
borneo2$pc2 <- pc8[["scores"]][,2]
borneo2$hull <- borneo2$Subspecies

autoplot(pc8, x = 1, y = 2, data = borneo2, colour = 'Subspecies', label = FALSE, size = 2.5, label.size = 3, frame=TRUE) + ggtitle("Mainland Bornean T. m. caedis vs. T. m. minor manus PCA")

find_hull <- function(borneo2) borneo2[chull(borneo2$pc1, borneo2$pc2), ]
hulls <- ddply(borneo2, "hull", find_hull)

m <- c("grey40", "grey80")
n <- c("grey80", "black")

a2 <-ggplot(data = borneo2, aes(x=pc1, y=pc2, group=Subspecies)) + scale_shape_manual(values = c(22,18), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + ylim(-3.1,3.5) + xlab("PC1 (46.05%)") + ylab("PC2 (22.06%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank(), legend.background = element_rect(fill = "transparent")) + theme(text = element_text(size=13)) + theme(legend.position = c(0.16, 0.95), legend.spacing.x = unit(2.5, 'mm'), legend.key.size = unit(2.5, 'mm'), legend.box.just = "left") + geom_polygon(data=hulls, alpha=0.3, aes(x=pc1, y=pc2, group=hull, fill=hull, linetype=hull), colour="black", show_guide = FALSE) + scale_linetype_manual(values=c(1,2)) + scale_fill_manual(values=m) + geom_point(aes(shape=Subspecies), color = "black", fill = "white", size=3) + annotate("text", x=4.1, y=3.2, label= "A", size = 10)

a2

ggsave(a2,file="Fig. 2A proofs.pdf",width=100, height=100, units="mm")
#ggsave(a2,file="Fig. 2A.eps",width=100, height=100, units="mm")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#skull
borneo <- x[x$Island == "Borneo",]

borneo <- na.omit(subset(borneo, select = c(Museum, Num, Subspecies, Sex, region, Type, Island, lat, long, UTL, PPL, LIB, MH, MCH, MCW, MCIL, LTL)))

borneo$Subspecies <- droplevels(borneo$Subspecies)
table(borneo$Subspecies)

pc7 <- princomp(borneo[,-c(1:9)], cor=TRUE)
for (i in 1:8) {
  pc7$loadings[,i] <- (pc7$loadings[,i] * pc7$sdev[i])
}
print(summary(pc7),digits=2,loadings=pc7$loadings,cutoff=0)
round(pc7$sdev^2,2)

borneo$pc1 <- pc7[["scores"]][,1]
borneo$pc2 <- pc7[["scores"]][,2]
borneo$hull <- borneo$Subspecies

autoplot(pc7, x = 1, y = 2, data = borneo, colour = 'Subspecies', shape = 'Sex', label = FALSE, size = 2.5, label.size = 3, frame=TRUE) + ggtitle("Mainland Bornean T. m. caedis vs. T. m. minor skull PCA")

find_hull2 <- function(borneo) borneo[chull(borneo$pc1, x1$pc2), ]
hulls2 <- ddply(borneo, "hull", find_hull)

b2 <-ggplot(data = borneo, aes(x=pc1, y=pc2, group=Subspecies)) + scale_shape_manual(values = c(22,18), guide = guide_legend(label.theme = element_text(angle = 0, face = "italic", size=9))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (54.25%)") + ylab("PC2 (20.31%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank()) + theme(text = element_text(size=13)) + theme(legend.position = c(0.16, 0.95), legend.spacing.x = unit(2.5, 'mm'), legend.key.size = unit(2.5, 'mm'), legend.box.just = "left", legend.background = element_rect(fill = "transparent")) + geom_polygon(data=hulls2, alpha=0.3, aes(x=pc1, y=pc2, group=hull, fill=hull, linetype=hull), colour="black", show_guide = FALSE) + scale_linetype_manual(values=c(1,2)) + scale_fill_manual(values=m) + geom_point(aes(shape=Subspecies), color = "black", fill = "white", size=3) + annotate("text", x=6.3, y=3, label= "B", size = 10) + xlim(-4.45,6.8) + ylim(-2.28,3.2)

b2

ggsave(b2,file="Fig. 2B proofs.pdf",width=100, height=100, units="mm")
```

###Manus PCA with all islands

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ytot <- na.omit(subset(y, select = c(Inst, Cat., Subspecies, Island, region, lat, long, Sex, Side, X1ML, X1PPL, X1MD, X1PPD, X2MW, X3DPL, X2PPW, X4PPW)))

ytot$Subspecies[ytot$Subspecies == "T. m. caedis"] <- "T. m. minor"

ytot$Island <- droplevels(ytot$Island)
table(ytot$Island)

u2 <- ytot[which(ytot$Island == "Pulo Laut" | ytot$Island == "Singkep" | ytot$Island == "Lingga"),]
u2$Island <- droplevels(u2$Island)

pc24 <- princomp(ytot[,-c(1:9)], cor=TRUE)

for (i in 1:8) {
  pc24$loadings[,i] <- (pc24$loadings[,i] * pc24$sdev[i])
}
print(summary(pc24),digits=2,loadings=pc24$loadings,cutoff=0)
round(pc24$sdev^2,2)

ytot$pc1 <- pc24[["scores"]][,1]
ytot$pc2 <- pc24[["scores"]][,2]
ytot$hull <- ytot$Subspecies

autoplot(pc24, x = 1, y = 2, data = ytot, colour = 'Island', label = FALSE, size = 2.5, label.size = 3, frame=TRUE) + ggtitle("All Islands, manus")

find_hull <- function(ytot) ytot[chull(ytot$pc1, ytot$pc2), ]
hulls3 <- ddply(ytot, "hull", find_hull)

n <- c("grey60", "grey30", "grey80", "black", "grey60")

a3 <- ggplot(data = ytot, aes(x=pc1, y=pc2, group=Island)) + scale_shape_manual(values = c(18,15,21,16,23,24), guide = guide_legend(label.theme = element_text(angle = 0, size=7))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (45.54%)") + ylab("PC2 (18.12%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank(), legend.background=element_rect(fill = "transparent")) + theme(text = element_text(size=10)) + theme(legend.position = c(0.002, 0.002), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(2, 'mm'), legend.justification = c("left", "bottom"),  legend.box.just = "left") + geom_polygon(data=hulls3, alpha=0.3, aes(x=pc1, y=pc2, group=hull, fill=hull, linetype=hull), colour="black", show_guide = FALSE) + scale_linetype_manual(values=c(5,1,10,2)) + scale_fill_manual(values=n) + geom_point(aes(shape=Island), color = "black", fill = "white", size=2.5) + annotate("text", x=5, y=2.5, label= "A", size = 10) + xlim(-4.9,5.5) + ylim(-3.7, 2.85) + theme(legend.key.size = unit(0.7, 'lines')) + geom_text(label="T. m. minor", x=-3.8, y=2.7, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="T. m.", x=-4.3, y=-1.15, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="malaccana", x=-4.3, y=-1.4, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="T. m. humeralis", x=4.3, y=-3.45, size = 3, color = "black", fontface = "italic", check_overlap = T)

a3

ggsave(a3,file="Fig. 3A proofs.pdf",width=100, height=100, units="mm")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ytot2 <- ytot[-which(ytot$Island == "Pulo Laut" | ytot$Island == "Singkep" | ytot$Island == "Lingga"),]

linfiny2 <- lda(ytot2[,-c(1:9)],grouping=ytot2$Island, CV=TRUE)
ctrawfiny2 <- table(ytot2$Island, linfiny2$class)
ctrawfiny2 #rows are actual count and columns are predicted count

# total percent correct: jackknife
round(sum(diag(prop.table(ctrawfiny2))),4)

linfiny2 <- lda(ytot2[,-c(1:9)],grouping=ytot2$Island, CV=FALSE)
ctrawfiny2 <- table(ytot2$Island, predict(linfiny2)$class)
ctrawfiny2 #rows are actual count and columns are predicted count

# total percent correct: no jackknife
round(sum(diag(prop.table(ctrawfiny2))),4)

predict(linfiny2, u2[,-c(1:9)])$class
predict(linfiny2, CV=TRUE, u2[,-c(1:9)])$class
```

DFA: 72.06%; Jackknife 60.29%

When the islands are excluded from the original training dataset and treated as unknowns, they are classified as follows:
Lingga (n = 2) classified as Borneo;
Pulo Laut (n = 1) classified as Borneo;
Singkep (n = 1) classified as Borneo

###Skull PCA with all islands

```{r, echo = FALSE, warning = FALSE, message = FALSE}
xtot <- na.omit(subset(x, select = c(Museum, Num, Subspecies, Sex, region, Type, Island, lat, long, UTL, PPL, LTPL, MH, MCH, MCW, MCIL, LTL)))

xtot$Subspecies[xtot$Subspecies == "T. m. caedis"] <- "T. m. minor"

xtot$Island <- droplevels(xtot$Island)
table(xtot$Island)

pc23 <- princomp(xtot[,-c(1:9)], cor=TRUE)
for (i in 1:8) {
  pc23$loadings[,i] <- (pc23$loadings[,i] * pc23$sdev[i])
}
print(summary(pc23),digits=2,loadings=pc23$loadings,cutoff=0)
round(pc23$sdev^2,2)

xtot$pc1 <- pc23[["scores"]][,1]
xtot$pc2 <- pc23[["scores"]][,2]
xtot$hull <- xtot$Subspecies

autoplot(pc23, x = 1, y = 2, data = xtot, colour = 'Island', label = FALSE, size = 2.5, label.size = 3, frame=TRUE) + ggtitle("All Islands, skull")

find_hull <- function(xtot) xtot[chull(xtot$pc1, xtot$pc2), ]
hulls4 <- ddply(xtot, "hull", find_hull)

n <- c("grey60", "grey30", "grey80", "black", "grey60")

b3 <- ggplot(data = xtot, aes(x=pc1, y=pc2, group=Island)) + scale_shape_manual(values = c(15,23,18,16,24), guide = guide_legend(label.theme = element_text(angle = 0, size=7.5))) + geom_vline(aes(xintercept = 0), size=0.25) + geom_hline(aes(yintercept = 0), size=0.25) + theme_bw() + xlab("PC1 (60.38%)") + ylab("PC2 (19.53%)") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + theme(legend.title = element_blank(), legend.background = element_rect(fill="transparent")) + theme(text = element_text(size=10)) + theme(legend.position = c(0.002, 0.002), legend.spacing.x = unit(2, 'mm'), legend.key.size = unit(2, 'mm'), legend.justification = c("left", "bottom"),  legend.box.just = "left") + geom_polygon(data=hulls4, alpha=0.3, aes(x=pc1, y=pc2, group=hull, fill=hull, linetype=hull), colour="black", show_guide = FALSE) + scale_linetype_manual(values=c(5,1,10,1)) + scale_fill_manual(values=n) + geom_point(aes(shape=Island), color = "black", fill = "white", size=2.5) + annotate("text", x=6, y=2.7, label= "B", size = 10) + xlim(-5,6.5) + ylim(-3.3, 3) + theme(legend.key.size = unit(0.65, 'lines')) + geom_text(label="T. m. minor", x=2.5, y=2.3, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="T. m.", x=-4.5, y=2.45, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="malaccana", x=-4.3, y=2.2, size = 3, color = "black", fontface = "italic", check_overlap = T) + geom_text(label="T. m. humeralis", x=4.65, y=-3, size = 3, color = "black", fontface = "italic", check_overlap = T)

ggsave(b3,file="Fig. 3B proofs.pdf",width=100, height=100, units="mm")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
u <- xtot[which(xtot$Island == "Banggi" | xtot$Island == "Balambangan"),]
xtot2 <- xtot[-which(xtot$Island == "Banggi" | xtot$Island == "Balambangan"),]

linfin2 <- lda(xtot2[,-c(1:9)],grouping=xtot2$Island, CV=FALSE)
ctrawfin2 <- table(xtot2$Island, predict(linfin2)$class)
ctrawfin2 #rows are actual count and columns are predicted count

# total percent correct
round(sum(diag(prop.table(ctrawfin2))),4)

predict(linfin2, u[,-c(1:9)])$class
predict(linfin2, CV=TRUE, u[,-c(1:9)])$class

linfin2 <- lda(xtot2[,-c(1:9)],grouping=xtot2$Island, CV=TRUE)
ctrawfin2 <- table(xtot2$Island, linfin2$class)
ctrawfin2 #rows are actual count and columns are predicted count

# total percent correct
round(sum(diag(prop.table(ctrawfin2))),4)
```

DFA: 78.69%; Jackknife 72.13%

When the islands are excluded from the original training dataset and treated as unknowns, they are classified as follows:
Balambangan (n = 5) classified as Borneo;
Banggi (n = 2) classified as Borneo.
